Six ways to understand monads\\ \small{Experiences from the functional programming course at TU Delft}

### 2012-10-30 latex/paper.tex
Choosing the programming language

\documentclass[conference]{IEEEtran}
\usepackage{graphicx}
\usepackage{times}
\usepackage{subfigure}

@@

\subsection{Heterogeneous Background of participants}

@@

The expected diverse backgrounds of the students guaranteed that no generic
introduction to the topic would be sufficient. While the

@@

@@

The exercise consisted of rendering the Utah Teapot~\cite{Torre06} using any
graphics primitive of the student's choice using only right triangles.  In its
core, the exercise required students to decompose arbitrary triangles, which
comprised the input Utah Teapot model, to a series right triangles. The
students had to come up with the decomposition method (using some form of
analytic geometry), a decomposition restriction criterion to stop the
decomposition when triangles could not be rendered on screen and a method to
recursively apply the above mentioned transformations on the input data. As
always the students could decide the implementation language of their choice.

\begin{figure*}
\centering
\subfigure[Reference, rendered with Java graphics]{
\includegraphics[scale=0.173]{teapot-reference.png}
\label{fig:teapot-reference}
}
\subfigure[Decomposition in Scala, rendered with Java graphics]{
\includegraphics[scale=0.17]{teapot-java-canvas.png}
\label{fig:teapot-scala}
}
\subfigure[Decomposition in Scheme, rendered with appropriately positioned {\sc html} \textsf{div}
elements]{
\includegraphics[scale=0.2]{teapot-html-divs.png}
\label{fig:teapot-html-divs}
}
\subfigure[Decomposition in Javascript, rendered with {\sc html5} graphics]{
\includegraphics[scale=0.31]{teapot-html-canvas.png}
\label{fig:teapot-html-canvas}
}
\label{fig:proj}
\caption[]{Example results from the teapot experiment.}
\end{figure*}

The students' response to the exercise was overwhelming. For four days,
the students worked hard to

@@

We would like to thank the course participants for their enthusiasm and
dedication.

>>>>

\documentclass[conference]{IEEEtran}
\usepackage{graphicx}
\usepackage{times}
\usepackage{subfigure}
\usepackage{url}

@@

\subsection{Heterogeneous background of participants}

@@

The expected diverse backgrounds of the students guaranteed that no generic
introduction to the topic would be sufficient. For this reason, we decided
to adopt a hands-on-first approach; the students would have to learn by
flexing their programming muscles, instead of being gently introduced to 
the theory through toy examples.

@@

Programming languages, in addition to enabling the programmer to express a
series of instructions to be executed by a computer, affect the programmer's
tough process and consequently her approach towards problem
solving~\cite{Ivers80}.  When teaching a programming course, it is important to
be able to demonstrate concepts without interference by the chosen language's
syntax. However, a non-practical language might have the opposite effect; our
experience has shown that the further a demonstration language is from practical
application, the less important student feel the taught concepts are.
Fortunately, most functional programming concepts can be expressed cleanly by
several widespread languages: for example, Javascript, Ruby and {\sc c\#} have
closures, first class and higher order functions, while a number of emerging
languages, such as Scala, {\sc f\#} and Clojure are functional languages working
on familiar development platforms. Apart from staying compatible with the
existing literature, there is no practical reason that necessitates teaching of
functional programming strictly in a language like Haskell or {\sc ml}.

The above led us to not choose any particular language for the course. 
During the lectures, the demonstration languages would be Haskell and
{\sc c\#}. We actively encouraged students to use the language of their
choice on the platform of their choice to carry out homework assignments
and the course's final project.

@@

The exercise consisted of rendering the Utah Teapot~\cite{Torre06} using any
graphics primitive of the student's choice using only right triangles.  In its
core, the exercise required students to decompose arbitrary triangles, which
comprised the input Utah Teapot model, to a series right triangles. The students
had to come up with the decomposition method (using some form of analytic
geometry), a decomposition restriction criterion to stop the decomposition when
triangles could not be rendered on screen and a method to recursively apply the
above mentioned transformations on the input data. As always, the students could
decide the implementation language of their choice. The exercise was given as a
mid-week homework assignment, between the Monday and the Friday lectures.

\begin{figure*}
\centering
\label{fig:teapots}
\subfigure[Input drawing, rendered with Java graphics]{
\includegraphics[scale=0.173]{teapot-reference.png}
\label{fig:teapot-reference}
}
\subfigure[Decomposition in Scala, rendered with Java graphics]{
\includegraphics[scale=0.17]{teapot-java-canvas.png}
\label{fig:teapot-scala}
}
\subfigure[Decomposition in Scheme, rendered with appropriately positioned {\sc html} \textsf{div}
elements]{
\includegraphics[scale=0.2]{teapot-html-divs.png}
\label{fig:teapot-html-divs}
}
\subfigure[Decomposition in Javascript, rendered with {\sc html5} graphics]{
\includegraphics[scale=0.31]{teapot-html-canvas.png}
\label{fig:teapot-html-canvas}
}
\caption[]{Example results from the teapot exercise.}
\end{figure*}

The student's response was overwhelming. Even though it was made clear that
the exercise would not contribute to the final grade, the will to apply the
data processing techniques taught during the lectures motivated
the students to work very hard. The fact that they were instructed to use
their favourite language enabled them to focus on decomposing the problem 
in a series of testable calculation steps rather than meddling with the
intricacies of a pure functional language. Solutions where provided in
Scala, {\sc c\#}, Scheme, Javascript and Haskell. 
Example renderings produced by student programs can be seen in 
Figure~\ref{fig:teapots}.

@@

We would like to thank the course participants for their enthusiasm and
dedication. The source code to all student projects and the teapot
exercise can be found at (or linked from) \url{https://github.com/gousiosg/teapots}.
